{ \sffamily \LARGE Electronic supplementary material}\\{\bf \sffamily \LARGE Evolution of reciprocity with limited payoff memory

### 2024-03-20 paper/main.tex
fix keyword order
submit revisions

~\\
{\it Keywords:} Evolution of cooperation; evolutionary dynamics; direct reciprocity; repeated prisoner's dilemma; social learning

@@

Evolutionary game theory describes the dynamics of populations when an individual's fitness depends on the traits or strategies of other population members~\cite{hofbauer1998evolutionary, nowak:Nature:2004, hauert2005game,Traulsen:PhilTrans:2022}.  
This theory can be used to describe the dynamics of animal conflict~\citep{maynard-smith:Nature:1973}, cancer cells~\citep{Stein:PTRS:2023}, and of cooperation~\citep{nowak:Science:2006}. 
Respective models translate strategic interactions into games~\cite{smith1982evolution}. 
These games specify how individuals (players) interact, which strategies individuals can choose, and what fitness consequences (or payoffs) the different strategies have. 
In addition, these models also specify the mode by which successful strategies spread over time. 
In models of biological evolution, individuals with a high fitness produce more offspring; in models of cultural evolution, such individuals are imitated more often. 
Although biological and cultural evolution are sometimes treated as equivalent, there can be important differences~\citep{Wu2015,Smolla:PTRS:2021,Denton:TPB:2022}. 
For example, models of biological evolution do not require individuals to have any particular cognitive abilities.
Here, it is the evolutionary process itself that biases the population towards strategies with higher fitness. 
In contrast, in models of cultural evolution, individuals need to be aware of the different strategies present in the population, and they need to identify those strategies with a higher payoff. 
As a consequence, evolutionary outcomes may depend on how easily different behaviors can be learned~\citep{Chatterjee:JTB:2012}, as well as on how readily payoffs can be compared.

@@

Most models of direct reciprocity do not address these difficulties~\citep{brauchli:JTB:1999,brandt:JTB:2006,ohtsuki:JTB:2007b,szolnoki:pre:2009b,imhof2010stochastic,van-segbroeck:prl:2012,grujic:jtb:2012,Martinez2012,stewart:pnas:2013,pinheiro:PLoSCB:2014,stewart:games:2015,Baek2016,McAvoy:ProcA:2019,glynatsi:SCR:2020,Schmid:PlosCB:2022,Murase:SciRep:2022,Cooney:BMB:2022,Chen:PNASNexus:2023} and, instead, assume individuals can easily copy the strategies of others. 
Similarly, they assume that updating decisions are based on the strategies' average (or expected) payoffs, which are based on all rounds and all interactions. 
These assumptions create a curious inconsistency in how models represent an individual's cognitive abilities. 
On the one hand, when playing the game, individuals are often assumed to have restricted memory. 
Respective studies typically assume that individuals make their decisions each round based on the outcome of the last round only~\citep[with only a few exceptions, see Refs.][]{Hauert1997,van-veelen:PNAS:2012,Stewart2016,hilbe:PNAS:2017,Li:NatCS:2022,Murase:PLoSCompBio:2023a}. 
Yet when learning new strategies, individuals are assumed to remember (or compute) each others' precise average payoff across many rounds and many interaction partners. 
Herein, we wish to explore whether this latter assumption is actually necessary for the evolution of reciprocity through social imitation. 
We ask whether individuals can learn to adopt reciprocal strategies even when learning is based on payoff information from a limited number of rounds.

@@

%% RESULTS: Snowdrift Game
In a second step, we also analyze the evolutionary dynamics of a different social dilemma, the snowdrift game~\cite{doebeli:Ecology:2005}. 
In this game, individuals can again cooperate or defect. 
However, now they get the benefit $b$ if at least one of them cooperates. 
Cooperators in turn pay a cost of $c$ or $c/2$, depending on whether they are the only one to cooperate or not. 
Compared to the donation game, the snowdrift game represents a weaker form of a social dilemma, because defection is no longer the dominant action in any single round~\citep{nowak:JTB:2012}.
Again, we derive analytical results in the case of large populations and strong selection. 
For perfect memory, cooperation is stochastically stable when \(q < 1 \!-\! c/(2 \delta b)\). 
Interestingly, for limited payoff memory, we obtain the same condition as for the donation game, \(q < 1 \!-\! 1/(2 \delta)\). 
We explain this invariance in detail in the \esm{}: compared to the donation game, the ordering of one-shot payoffs in the snowdrift game only differs in one instance (the {\it sucker's payoff} now exceeds the {\it punishment payoff}); however, this instance is irrelevant for the stochastic stability of cooperation. 
Further evolutionary simulations suggest that again, perfect memory leads to higher cooperation rates than limited memory; but now even limited memory can result in substantial cooperation (Supplementary Figures~8, 9).

@@

Our main contribution is an application of these ideas to direct reciprocity, which is one of the key mechanisms to explain why unrelated individuals might cooperate~\citep{nowak:Science:2006}. 
According to this mechanism, cooperation pays if it makes the interaction partner more cooperative in future. 
To describe which strategies are most effective, the previous theoretical literature assumes that the evolutionary dynamics are driven by the players' expected payoffs~\citep{brauchli:JTB:1999,brandt:JTB:2006,ohtsuki:JTB:2007b,szolnoki:pre:2009b,imhof2010stochastic,van-segbroeck:prl:2012,grujic:jtb:2012,Martinez2012,stewart:pnas:2013,pinheiro:PLoSCB:2014,stewart:games:2015,Baek2016,McAvoy:ProcA:2019,glynatsi:SCR:2020,Schmid:PlosCB:2022,Murase:SciRep:2022}.
To the extent that strategies are learned (not inherited), this assumption seems to impose rather stringent requirements on the individuals' cognitive abilities. 
In the most extreme case, individuals would have to remember (or compute) their payoffs against all population members, for all possible ways in which their repeated games may unfold. 
This assumption introduces a curious inconsistency in how these models represent an individual's cognitive abilities. 
For playing their games, individuals are often assumed to only recall the outcome of the very last round. 
Yet to update their strategies, individuals are implicitly assumed to have a record of the outcome of all rounds, across all interaction partners.

@@

To explore the effect of payoff memory, we only considered the simplest model of reciprocity. 
This leaves several possible directions for future work. 
For example, for both our analytical analysis and our simulations, we assumed that populations are {\it well-mixed}. 
As a result, any population member is equally likely to be a given player's last co-player. 
A similar analysis for structured populations would be desirable, as population structure can often promote cooperation without an explicit need for reciprocity~\citep{szabo:PR:2007,van-veelen:PNAS:2012}.
Similarly, we assumed that after identifying a suitable role model, players can easily imitate that role model's strategy. 
In the context of our study, this might be a particular simplification, because players with limited payoff memory may also have less accurate information to infer a co-player's strategy. 
 Finally, we have taken a player's payoff memory to be a given parameter of the model. 
 Instead, future models could explore how that payoff memory might co-evolve along with the players' strategies \citep[similar to previous work on the evolution of the players' feasible strategy sets, e.g.][]{Stewart2016,Schmid:PlosCB:2022,Murase:PLoSCompBio:2023a}. 
 Here, individuals might experience an interesting trade-off. 
 Better payoff-memory would help them to make better decisions, yet it might also come with higher cognitive costs. 
 For such a setup, our results suggest that in many instances, a modest amount of memory may be sufficient to achieve good outcomes.

@@

\noindent
{\bf Data accessibility.}
All data and code used in this manuscript are openly available at\newline
\texttt{https://zenodo.org/records/10066227}.\\

@@

{
{\setlength{\bibsep}{0\baselineskip}
\bibliographystyle{naturemag}
\bibliography{bibliography}
}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{static/donation_expected_last_round_summary_results.pdf}
    \caption{{\bf Evolutionary dynamics under perfect and limited payoff memory.}
    The leftmost panels give a schematic overview of the two main scenarios we compare. 
    The two scenarios differ in how many past interactions individuals take into account when updating their strategy. 
    In the scenario with perfect payoff memory, individuals consider all their past interactions (against all population members, and taking every turn of each repeated game into account). 
    In the scenario with limited payoff memory, individuals only consider their very last interaction (against one specific population member, taking into account only one round of the repeated game). 
    The four panels  on the right side depict the outcome of evolutionary simulations for repeated games with either a low or a high benefit of cooperation. 
    Colors represent how often the respective region of the strategy space is visited over time. 
    In all four panels, two regions are visited particularly often. 
    One region corresponds to a neighborhood of \alld{} with $p\approx q\!\approx\!0$ (lower left corner). 
    The other region corresponds to a strip of conditionally cooperative strategies with $p\!\approx\! 1$ and $q$ satisfying the constraints \eqref{Eq:PerfectMemory} and \eqref{Eq:LimitedMemory}, respectively (lower right corner). 
    The resulting average cooperation rate depends on which of these two neighborhoods is visited more often.
    Simulations are run for $T\!=\!10^7$ time steps, using a cost $c\!=\!1$, a continuation probability of $\delta\!=\!0.999$ and a selection strength of $\beta\!=\!1$, in a population of size $N\!=\!100$.}
\label{fig:expected_and_stochastic_for_donation}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=.75\textwidth]{static/cooperation_rate_over_b_and_beta.pdf}
    \caption{{\bf Evolution of direct reciprocity for different parameter values.} 
    To explore the robustness of our results, we have run simulations for different benefit values (left panels, {\it a}), and for different selection strengths (right panels, {\it b}). 
    In each case, we record the resulting average cooperation rate over the entire simulation (upper panels). 
    In addition we record the individuals' average generosity. 
    Here, we only take into account those residents with $p\! \approx\! 1$ and we compute
    the average of their cooperation probability~$q$. 
    These simulations suggest that perfect payoff memory consistently leads to more cooperation and more generosity. 
    Unless explicitly varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$.
    Simulations are run for $T\!=\!5\times 10^7$ time steps, and each point represents a single simulation run.}
    \label{fig:cooperation_rate_over_benefit_and_beta}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{static/more_memory_summary_results.pdf}
  \caption{{\bf Average cooperation rates for different payoff memories.}
  We vary how much information individuals take into account when updating their strategies. 
  From left to right, we consider the following cases.
  ({\it i}) Updating occurs based on expected payoffs (perfect memory), 
  ({\it ii}) it occurs based on the last round of one interaction (limited memory), 
  ({\it iii}) based on the last round of two interactions,
  ({\it iv}) based on the last two rounds of one interaction,
  ({\it v}) based on the last two rounds of two interactions, and
  ({\it vi}) based on the average payoff of one interaction.
  Again, simulations are run either for a comparably low benefit of cooperation~($b/c\!=\!3$), or for a high benefit~($b/c\!=\!10$). 
  We observe that perfect memory always yields the highest cooperation rate. 
  However, when individuals take into account at least two past interactions -- cases ({\it iii}) to ({\it vi}) -- evolving cooperation rates are close to this optimum. 
  Simulations are run for $T\!=\!10^7$ time steps, based on the parameters $N\!=\!100$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
\label{fig:cooperation_rate_all_updating_payoffs}
\end{figure}

>>>>

~\\
{\it Keywords:} Direct reciprocity; evolution of cooperation; evolutionary dynamics; repeated prisoner's dilemma; social learning

@@

Evolutionary game theory describes the dynamics of populations when an individual's fitness depends on the traits or strategies of other population members~\citep{hofbauer1998evolutionary, nowak:Nature:2004, hauert2005game,Traulsen:PhilTrans:2022}.  
This theory can be used to describe the dynamics of animal conflict~\citep{maynard-smith:Nature:1973}, cancer cells~\citep{Stein:PTRS:2023}, and of cooperation~\citep{nowak:Science:2006}. 
Respective models translate strategic interactions into games~\citep{smith1982evolution}. 
These games specify how individuals (players) interact, which strategies individuals can choose, and what fitness consequences (or payoffs) the different strategies have. 
In addition, these models also specify the mode by which successful strategies spread over time. 
In models of biological evolution, individuals with a high fitness produce more offspring; in models of cultural evolution, such individuals are imitated more often. 
Although biological and cultural evolution are sometimes treated as equivalent, there can be important differences~\citep{Wu2015,Smolla:PTRS:2021,Denton:TPB:2022}. 
For example, models of biological evolution do not require individuals to have any particular cognitive abilities.
Here, it is the evolutionary process itself that biases the population towards strategies with higher fitness. 
In contrast, in models of cultural evolution, individuals need to be aware of the different strategies present in the population, and they need to identify those strategies with a higher payoff. 
As a consequence, evolutionary outcomes may depend on how easily different behaviors can be learned~\citep{Chatterjee:JTB:2012}, as well as on how readily payoffs can be compared.

@@

Most models of direct reciprocity do not address these difficulties~\citep{brauchli:JTB:1999,brandt:JTB:2006,ohtsuki:JTB:2007b,szolnoki:pre:2009b,imhof2010stochastic,vansegbroeck:prl:2012,grujic:jtb:2012,Martinez2012,stewart:pnas:2013,pinheiro:PLoSCB:2014,stewart:games:2015,Baek2016,McAvoy:ProcA:2019,glynatsi:SCR:2020,Schmid:PlosCB:2022,Murase:SciRep:2022,Cooney:BMB:2022,Chen:PNASNexus:2023} and, instead, assume individuals can easily copy the strategies of others. 
Similarly, they assume that updating decisions are based on the strategies' average (or expected) payoffs, which are based on all rounds and all interactions. 
These assumptions create a curious inconsistency in how models represent an individual's cognitive abilities. 
On the one hand, when playing the game, individuals are often assumed to have restricted memory. 
Respective studies typically assume that individuals make their decisions each round based on the outcome of the last round only~\citep[with only a few exceptions, see Refs.]{Hauert1997,vanveelen:PNAS:2012,Stewart2016,hilbe:PNAS:2017,Li:NatCS:2022,Murase:PLoSCompBio:2023a}. 
Yet when learning new strategies, individuals are assumed to remember (or compute) each others' precise average payoff across many rounds and many interaction partners. 
Herein, we wish to explore whether this latter assumption is actually necessary for the evolution of reciprocity through social imitation. 
We ask whether individuals can learn to adopt reciprocal strategies even when learning is based on payoff information from a limited number of rounds.

@@

%% RESULTS: Snowdrift Game
In a second step, we also analyze the evolutionary dynamics of a different social dilemma, the snowdrift game~\citep{doebeli:Ecology:2005}. 
In this game, individuals can again cooperate or defect. 
However, now they get the benefit $b$ if at least one of them cooperates. 
Cooperators in turn pay a cost of $c$ or $c/2$, depending on whether they are the only one to cooperate or not. 
Compared to the donation game, the snowdrift game represents a weaker form of a social dilemma, because defection is no longer the dominant action in any single round~\citep{nowak:JTB:2012}.
Again, we derive analytical results in the case of large populations and strong selection. 
For perfect memory, cooperation is stochastically stable when \(q < 1 \!-\! c/(2 \delta b)\). 
Interestingly, for limited payoff memory, we obtain the same condition as for the donation game, \(q < 1 \!-\! 1/(2 \delta)\). 
We explain this invariance in detail in the \esm{}: compared to the donation game, the ordering of one-shot payoffs in the snowdrift game only differs in one instance (the {\it sucker's payoff} now exceeds the {\it punishment payoff}); however, this instance is irrelevant for the stochastic stability of cooperation. 
Further evolutionary simulations suggest that again, perfect memory leads to higher cooperation rates than limited memory; but now even limited memory can result in substantial cooperation (Supplementary Figures~8, 9).

@@

Our main contribution is an application of these ideas to direct reciprocity, which is one of the key mechanisms to explain why unrelated individuals might cooperate~\citep{nowak:Science:2006}. 
According to this mechanism, cooperation pays if it makes the interaction partner more cooperative in future. 
To describe which strategies are most effective, the previous theoretical literature assumes that the evolutionary dynamics are driven by the players' expected payoffs~\citep{brauchli:JTB:1999,brandt:JTB:2006,ohtsuki:JTB:2007b,szolnoki:pre:2009b,imhof2010stochastic,vansegbroeck:prl:2012,grujic:jtb:2012,Martinez2012,stewart:pnas:2013,pinheiro:PLoSCB:2014,stewart:games:2015,Baek2016,McAvoy:ProcA:2019,glynatsi:SCR:2020,Schmid:PlosCB:2022,Murase:SciRep:2022}.
To the extent that strategies are learned (not inherited), this assumption seems to impose rather stringent requirements on the individuals' cognitive abilities. 
In the most extreme case, individuals would have to remember (or compute) their payoffs against all population members, for all possible ways in which their repeated games may unfold. 
This assumption introduces a curious inconsistency in how these models represent an individual's cognitive abilities. 
For playing their games, individuals are often assumed to only recall the outcome of the very last round. 
Yet to update their strategies, individuals are implicitly assumed to have a record of the outcome of all rounds, across all interaction partners.

@@

To explore the effect of payoff memory, we only considered the simplest model of reciprocity. 
This leaves several possible directions for future work. 
For example, for both our analytical analysis and our simulations, we assumed that populations are {\it well-mixed}. 
As a result, any population member is equally likely to be a given player's last co-player. 
A similar analysis for structured populations would be desirable, as population structure can often promote cooperation without an explicit need for reciprocity~\citep{szabo:PR:2007,vanveelen:PNAS:2012}.
Similarly, we assumed that after identifying a suitable role model, players can easily imitate that role model's strategy. 
In the context of our study, this might be a particular simplification, because players with limited payoff memory may also have less accurate information to infer a co-player's strategy. 
 Finally, we have taken a player's payoff memory to be a given parameter of the model. 
 Instead, future models could explore how that payoff memory might co-evolve along with the players' strategies \citep[similar to previous work on the evolution of the players' feasible strategy sets, e.g.][]{Stewart2016,Schmid:PlosCB:2022,Murase:PLoSCompBio:2023a}. 
 Here, individuals might experience an interesting trade-off. 
 Better payoff-memory would help them to make better decisions, yet it might also come with higher cognitive costs. 
 For such a setup, our results suggest that in many instances, a modest amount of memory may be sufficient to achieve good outcomes.

@@

\noindent
{\bf Data accessibility.}
All data generated for this manuscript are openly available at:
\href{https://zenodo.org/records/10844104}{https://zenodo.org/records/10844104}.
The code for producing the results and generating the figures is available
on the online repository
\href{https://github.com/Nikoleta-v3/evolution-of-cooperation-with-limited-updating-payoff-memory}{evolution-of-cooperation-with-limited-updating-payoff-memory}.\\

@@

{
{\setlength{\bibsep}{0\baselineskip}
\bibliographystyle{naturemag}
\bibliography{bibliography}
}
}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{donation_expected_last_round_summary_results.pdf}
    \caption{{\bf Evolutionary dynamics under perfect and limited payoff memory.}
    The leftmost panels give a schematic overview of the two main scenarios we compare. 
    The two scenarios differ in how many past interactions individuals take into account when updating their strategy. 
    In the scenario with perfect payoff memory, individuals consider all their past interactions (against all population members, and taking every turn of each repeated game into account). 
    In the scenario with limited payoff memory, individuals only consider their very last interaction (against one specific population member, taking into account only one round of the repeated game). 
    The four panels  on the right side depict the outcome of evolutionary simulations for repeated games with either a low or a high benefit of cooperation. 
    Colors represent how often the respective region of the strategy space is visited over time. 
    In all four panels, two regions are visited particularly often. 
    One region corresponds to a neighborhood of \alld{} with $p\approx q\!\approx\!0$ (lower left corner). 
    The other region corresponds to a strip of conditionally cooperative strategies with $p\!\approx\! 1$ and $q$ satisfying the constraints \eqref{Eq:PerfectMemory} and \eqref{Eq:LimitedMemory}, respectively (lower right corner). 
    The resulting average cooperation rate depends on which of these two neighborhoods is visited more often.
    Simulations are run for $T\!=\!10^7$ time steps, using a cost $c\!=\!1$, a continuation probability of $\delta\!=\!0.999$ and a selection strength of $\beta\!=\!1$, in a population of size $N\!=\!100$.}
\label{fig:expected_and_stochastic_for_donation}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=.75\textwidth]{cooperation_rate_over_b_and_beta.pdf}
    \caption{{\bf Evolution of direct reciprocity for different parameter values.} 
    To explore the robustness of our results, we have run simulations for different benefit values (left panels, {\it a}), and for different selection strengths (right panels, {\it b}). 
    In each case, we record the resulting average cooperation rate over the entire simulation (upper panels). 
    In addition we record the individuals' average generosity. 
    Here, we only take into account those residents with $p\! \approx\! 1$ and we compute
    the average of their cooperation probability~$q$. 
    These simulations suggest that perfect payoff memory consistently leads to more cooperation and more generosity. 
    Unless explicitly varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$.
    Simulations are run for $T\!=\!5\times 10^7$ time steps, and each point represents a single simulation run.}
    \label{fig:cooperation_rate_over_benefit_and_beta}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{more_memory_summary_results.pdf}
  \caption{{\bf Average cooperation rates for different payoff memories.}
  We vary how much information individuals take into account when updating their strategies. 
  From left to right, we consider the following cases.
  ({\it i}) Updating occurs based on expected payoffs (perfect memory), 
  ({\it ii}) it occurs based on the last round of one interaction (limited memory), 
  ({\it iii}) based on the last round of two interactions,
  ({\it iv}) based on the last two rounds of one interaction,
  ({\it v}) based on the last two rounds of two interactions, and
  ({\it vi}) based on the average payoff of one interaction.
  Again, simulations are run either for a comparably low benefit of cooperation~($b/c\!=\!3$), or for a high benefit~($b/c\!=\!10$). 
  We observe that perfect memory always yields the highest cooperation rate. 
  However, when individuals take into account at least two past interactions -- cases ({\it iii}) to ({\it vi}) -- evolving cooperation rates are close to this optimum. 
  Simulations are run for $T\!=\!10^7$ time steps, based on the parameters $N\!=\!100$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
\label{fig:cooperation_rate_all_updating_payoffs}
\end{figure}
